---
title: "virus_project: initial batch of 4 files"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reading files

First, read four virus files:
- picornaviridae_aligned
- potyviridae_aligned
- picornaviridae_5UTR_aligned, and 
- potyviridae_5UTR_aligned 

into dataframes, which are prepared for later ancestral construction.


```{r}
library(ape)
library(phyloch) 
library(phylotools)

ReadFasta<-function(file) {
  # Read the file line by line
  fasta<-readLines(file)
  # Identify header lines
  ind<-grep(">", fasta)
  # Identify the sequence lines
  s<-data.frame(ind=ind, from=ind+1, to=c((ind-1)[-1], length(fasta)))
  # Process sequence lines
  seqs<-rep(NA, length(ind))
  for(i in 1:length(ind)) {
    seqs[i]<-paste(fasta[s$from[i]:s$to[i]], collapse="")
  }
  # Create a data frame 
  DF<-data.frame(name=gsub(">", "", fasta[ind]), sequence=seqs)
  # Return the data frame as a result object from the function
  return(DF)
}

# The sequences of picornaviridae, 226 virus
seq_picornaviridae=ReadFasta("picornaviridae_aligned.fasta")
nrow(seq_picornaviridae)

# The sequences of potyviridae, 218 virus
seq_potyviridae=ReadFasta("potyviridae_aligned.fasta")
nrow(seq_potyviridae)

# The sequences of picornaviridae_5UTR, 129 virus
seq_picornaviridae_5UTR=ReadFasta("picornaviridae_5UTR_aligned.fas")
nrow(seq_picornaviridae_5UTR)

# The sequences of potyviridae_5UTR, 152 virus
seq_potyviridae_5UTR=ReadFasta("potyviridae_5UTR_aligned.fas")
nrow(seq_potyviridae_5UTR)

```

### RAxML runs

Second, use Raxml ( 20 ML search + bootstrapping ) to get the tree with best likelihood. This part of commands is from bash (Terminal).

Note1: the file size of picornaviridae_aligned and potyviridae_aligned is too large. As a result, I limit the bootstrap search time to 1. The search times for Maximal Likelihood (ML) are always 20 times. The estimated best tree could be further improved if the bootstrap times increase. The bootstrap search times for picornaviridae_5UTR_aligned and potyviridae_5UTR_aligned are 6 and 10 times.

Note2: For picornaviridae_5UTR_aligned and potyviridae_5UTR_aligned, the names of some virus sequences contain colon (:), so I tried to delete the colon when the virus name contain it in order to make the fas file executable in Raxml. The files with deleted colon are picornaviridae_5UTR_aligned1 and potyviridae_5UTR_aligned1. 

In the picornaviridae_5UTR_aligned, names of five virus sequences contain colon (:). They are 
- NC_004421.1_Aichivirus_B_genomic_RNA_complete_genome_strain_U-1,NC_033695.1_Enterovirus_AN12_genomic_RNA_complete_genome_strain_AN12,
- NC_027919.1_Kobuvirus_cattle/Kagoshima-1-22-KoV/2014/JPN_genomic_RNA_nearly_complete_genome_strain_Kagoshima-1-22-KoV/2014/JPN,
- NC_027918.1_Kobuvirus_cattle/Kagoshima-2-24-KoV/2015/JPN_genomic_RNA_complete_genome_strain_Kagoshima-2-24-KoV/2015/JPN,
- NC_025474.1_Crohivirus_gene_for_polyprotein_complete_cds_strain_ZM54. 

For potyviridae_5UTR_aligned, the names of virus sequences contain colon (:) are  
- NC_014327.1_Pepper_yellow_mosaic_virus_RNA_complete_genome_strain_DF,NC_030391.1_Wild_onion_symptomless_virus_gene_for_polyprotein_complete_cds_isolate_TUR256-1,
- NC_021786.1_Habenaria_mosaic_virus_genomic_RNA_complete_genome_isolate_Ha-1

### Bash code for rapid bootstrapping in Raxml: 
```
raxmlHPC -f a -m GTRGAMMA -p 12345 -x 12345 -# 1 -s picornaviridae_aligned.fasta -n T1      

raxmlHPC -f a -m GTRGAMMA -p 12345 -x 12345 -# 1 -s potyviridae_aligned.fasta -n T2 

raxmlHPC -f a -m GTRGAMMA -p 12345 -x 12345 -#6 -s picornaviridae_5UTR_aligned1.fas -n T3 

raxmlHPC -f a -m GTRGAMMA -p 12345 -x 12345 -# 10 -s potyviridae_5UTR_aligned1.fas -n T4 
```

#### Relevant parameters for each ML search + bootstrapping best tree estimation

picornaviridae_aligned.fasta: alpha: 4.934338, Tree-Length: 126.910679, rate A <-> C: 1.448674, rate A <-> G: 3.216025, rate A <-> T: 1.314619, rate C <-> G: 1.269701, rate C <-> T: 3.239043, rate G <-> T: 1.000000, pi(A,C,G,T) = c(0.270869, 0.221802, 0.224860, 0.282469).

potyviridae_aligned.fasta: alpha: 0.932693, Tree-Length: 132.828870, rate A <-> C: 2.486990, rate A <-> G: 3.739170, rate A <-> T: 1.369314, rate C <-> G: 1.513431, rate C <-> T: 5.417214, rate G <-> T: 1.000000, pi(A,C,G,T) =  c(0.313614,0.195751,0.233819,0.256815)

picornaviridae_5UTR_aligned.fas: alpha: 5.094113, Tree-Length: 110.802014, rate A <-> C: 1.172756, rate A <-> G: 2.585556, rate A <-> T: 1.493185, rate C <-> G: 0.866442, rate C <-> T: 2.625791, rate G <-> T: 1.000000, pi(A,C,G,T) = c(0.225632,0.257601,0.242809,0.273959)

potyviridae_5UTR_aligned.fas: alpha: 1.039252, Tree-Length: 115.318660, rate A <-> C: 1.037454, rate A <-> G: 1.791470, rate A <-> T: 1.154272, rate C <-> G: 0.616361, rate C <-> T: 2.463065, rate G <-> T: 1.000000, pi(A,C,G,T) = c(0.354004,0.236219,0.145031,0.264746)


### Estimate ancestral sequence

#### Read trees in R
Read the best tree into R for each virus file. ( Use `plot( )` to check. Display tree better with Dendroscope because there are lot of taxa)

```{r}

tree_string_picornaviridae = read.tree("RAxML_bestTree.T1")
#plot(tree_string_picornaviridae)

tree_string_potyviridae = read.tree("RAxML_bestTree.T2")
#plot(tree_string_potyviridae)

tree_string_picornaviridae_5UTR=read.tree("RAxML_bestTree.T3")
#plot(tree_string_picornaviridae_5UTR)

tree_string_potyviridae_5UTR=read.tree("RAxML_bestTree.T4")
#plot(tree_string_potyviridae_5UTR)

write.tree(tree_string_picornaviridae_5UTR,file="tree_3.newick")

```

#### Root trees in R (midpoint) 
Next, root the unrooted trees prepared for later ancestral construction. Midpoint rooting locates the midpoint of the longest path between any two tips and putting the root in that location. After using midpoint root on the tree string, `write.tree( )` function in R could help check how new rooted trees look like.

```{r}
library(phytools)

tree_string_picornaviridae_rooted=midpoint.root(tree_string_picornaviridae)

tree_string_potyviridae_rooted=midpoint.root(tree_string_potyviridae)

tree_string_picornaviridae_5UTR_rooted=midpoint.root(tree_string_picornaviridae_5UTR)

tree_string_potyviridae_5UTR_rooted=midpoint.root(tree_string_potyviridae_5UTR)

#write.tree(tree_string_picornaviridae_5UTR_rooted,file="tree5.newick")

```

#### Ancestral state reconstruction
```{r}
library(phytools)
library(magrittr)
library(kableExtra)
tree=read.tree("RAxML_bestTree.T3")
tree=midpoint.root(tree)
try1=fastBM(tree_string_picornaviridae_5UTR_rooted)
# The fitted paramter of reconstruction
fit<-fastAnc(tree_string_picornaviridae_5UTR_rooted,try1,vars=TRUE,CI=TRUE)

x<-fastBM(tree,internal=TRUE)
## ancestral states
a<-x[as.character(1:tree$Nnode+Ntip(tree))]
## tip data
x<-x[tree$tip.label]
fit<-fastAnc(tree,x,CI=TRUE)

plot(a,fit$ace,xlab="true states",ylab="estimated states")
phenogram(tree,x,ftype="off",spread.labels=FALSE)

phenogram(tree_string_picornaviridae_5UTR_rooted,try1,fsize=0.6,spread.costs=c(1,0))

```

Claudia question: Where is the file `tree3_N0-ancestors_GRASP.fasta` created?

# Show the ancestral construction using grasp

```{r}
#ancestral3=read.dna("tree3_N0-ancestors_GRASP.fasta",as.character=FALSE,format="fasta")
#ancestral3
#ancestral3_character=read.dna("tree3_N0-ancestors_GRASP.fasta",as.character=TRUE,format="fasta")

#ancestral4=read.dna("tree4_N0_ancestors_GRASP.fasta",as.character=FALSE,format="fasta")
#ancestral4
#ancestral4_character=read.dna("tree4_N0_ancestors_GRASP.fasta",as.character=TRUE,format="fasta")
```

# Output files

In `results/initial-files` folder:
- RAxML_bestTree.T1
- RAxML_bestTree.T2
- RAxML_bestTree.T3
- RAxML_bestTree.T4
- tree3_N0-ancestors_GRASP.fasta